Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-02-06 appendix.tex
comments
:+1: for paper reference
references, spacing, cover page

\subsection*{Site split formulation}
We begin by introducing ``site splits,'' which formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.

@@

We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

%EM You dropped the below parenthetical reminder. Did you think that it was clear?

@@

We use basin-hopping \cite{Wales1997} implemented in \texttt{scipy} \cite{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}$ as well as any optimization method available.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01-marginal.svg --out-pkl-name figures/w-hat-empirical-01-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01-marginal.svg --in-pkl-name figures/w-hat-empirical-01-marginal.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01-marginal}
\caption{
    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
}
\label{fig:bl-general-marginal}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
}
\label{fig:bl-general-bias}
\end{figure}

>>>>

\subsection*{Site split formulation}
We begin by introducing ``site splits,'' which formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.

@@

We state the Hadamard representation of site split generating probabilities, following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

@@

@@

We use basin-hopping \citep{Wales1997} implemented in \texttt{scipy} \citep{Jones2001} to obtain the plots in Figs.~\ref{fig:bl-general-inconsistency}, \ref{fig:bl-general-marginal}, and \ref{fig:bl-general-bias}.
This method randomly perturbs candidate solutions and accepts or rejects them based on the nearby likelihood surface, eventually obtaining an estimate of the ``global'' optimum.
While it is not guaranteed to always obtain a global optimum---or even always converge---its implementation here leads us to believe that we can estimate $\hat{w}$ as well as any optimization method available.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01-marginal.svg --out-pkl-name figures/w-hat-empirical-01-marginal.pkl --n-jobs 16 --marginal
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01-marginal.svg --in-pkl-name figures/w-hat-empirical-01-marginal.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01-marginal}
\caption{
    Estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing the classical marginal likelihood \eqref{eq:marginal_likelihood} (rather than a joint optimization procedure).
}
\label{fig:bl-general-marginal}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
}
\label{fig:bl-general-bias}
\end{figure}

### 2018-02-06 joint_inf.tex
references, spacing, cover page

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}
\usepackage{palatino}
\usepackage{hyphenat}
\usepackage{thm-restate}
\usepackage{lineno}
\linenumbers

@@

\title{Joint maximum-likelihood on phylogenies and ancestral states is not consistent}
\author{David A. Shaw \and Frederick A. Matsen IV}

\begin{document}
\maketitle

@@

@@

Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
In a recent paper, \cite{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
This latter conditional optimization is simpler and more computationally efficient than optimizing the marginal likelihood.
But is it statistically consistent?

An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although the software described in \cite{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \cite{Semple2003-em}.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
This parameter quantifies the fidelity of transmission of the ancestral state across an edge \cite{Matsen2007-jq}.

@@

The standard phylogenetic likelihood approach on unrooted trees under the usual assumption of independence between sites is as follows.
For a topology $\tau$ and branch fidelities $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \cite{Felsenstein2004}.

@@

The alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \cite{RoyChoudhury2015-ta}.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be one.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name figures/w-hat-empirical-01.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01}
\caption{
    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
This inconsistency in the case of long branches concurs with some empirical findings in \cite{Sagulenko2017-jo}, namely their Figures~2 and 3.

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \cite{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \cite{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint optimization procedure.

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \cite{Felsenstein1978-rr}.
In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
Moreover, the inconsistency in \cite{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \cite{Siddall1998-hq}, though \cite{Swofford2001-hr} show that sequence length plays a major role in these issues.

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the general case is not unlike the restricted case.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \cite{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin and Joe Felsenstein for helpful discussions.

@@

\bibliographystyle{plain}
\bibliography{joint_inf}

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}
\usepackage{palatino}
\usepackage{hyphenat}
\usepackage{thm-restate}
\usepackage{lineno}
\linenumbers
\usepackage{natbib}
\usepackage{setspace}
\doublespacing
\usepackage{authblk}
\usepackage{fancyhdr}

@@

\fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }

\title{Joint maximum-likelihood of phylogenies and ancestral states is not consistent}

\author[1]{David A. Shaw}
\author[1]{Frederick A. Matsen IV\thanks{Corresponding author, email: \url{matsen@fredhutch.org}}}
\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
\date{}

\begin{document}

@@

\maketitle
\thispagestyle{fancy}

@@

Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out latent ancestral states at the internal nodes of the tree.
In a recent paper, \citet{Sagulenko2017-jo} suggest using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences on a fixed topology.
This is attractive from a computational perspective: such joint inference can proceed according to an iterative procedure in which ancestral sequences are first estimated and model parameters are optimized conditional on these estimates.
This latter conditional optimization is simpler and more computationally efficient than optimizing the marginal likelihood.
But is it statistically consistent?

An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``Farris zone'' \citep{Siddall1998-hq} tree, and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \citep{Semple2003-em}.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called ``fidelity.''
The probability of a substitution on a branch with fidelity $\theta$ is $(1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$ where $0 \le \theta \le 1$.
This parameter quantifies the fidelity of transmission of the ancestral state across an edge \citep{Matsen2007-jq}.

@@

The standard phylogenetic likelihood approach on unrooted trees under the usual assumption of independence between sites is as follows.
For a topology $\tau$ and branch fidelities $t$ the likelihood given observed ancestral states $\fullAncestralStates$ is
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
The probability $\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)$ is a product of transition probabilities determined by $\fullAlignment$, $\fullAncestralStates$, $\tau$, and $t$ \citep{Felsenstein2004}.

@@

The alternative approach \citep{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
To determine how conservative, we use the method of basin-hopping \citep{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
This region encompasses the majority of the branch fidelity space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch fidelity to be one.

@@

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name figures/w-hat-empirical-01.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01}
\caption{
    Numerical estimates for $\hat{w}$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \citep{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint optimization procedure.

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the ``Felsenstein zone'' topology $\tau_2$, as the number of observations increases, the ``Farris zone'' topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the Farris topology.
Moreover, the inconsistency in \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
Difficulties in phylogenetic estimation when generating data on the ``Farris zone'' tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the general case is not unlike the restricted case.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.

@@

\bibliographystyle{plainnat}
\bibliography{joint_inf}
